Food-web structure, stochasticity and response diversity drive temporal community stability in complex communities

### 2024-04-12 paper/manuscript.Rmd
Rewrite intro and abstract in NEE style

- Putative titles:
  - Community stability is driven by biodiversity and response diversity in
    complex stochastic food-webs
  - Response diversity and community structure drive community stability in
    complex food-webs

@@

[NOT CHANGED] Empirical and theoretical evidence suggests environmental changes will
dramatically destabilise ecological communities. However, this evidence is
largely based on oversimplified communities, which lack the complexity of
natural communities. We characterise the dynamics of complex food-webs under a
stochastic food-web model, and investigate how response diversity,
stochasticity, and food-web structure drive temporal stability. We found that
higher average trophic level increased community stability by increasing
population stability, at the expense of asynchrony. Species richness had little
effect on population stability, but a large one on asynchrony. Interaction
strength and connectance had little effects on temporal stability overall, but
interacted strongly with response diversity to determine stability-richness
relationships. Absence of response diversity all leaded to negative
stability-richness relationships, but high response diversity resulted in high
prevalence of positive stability-richness relationships. Our study reveals that
response diversity and food-web structure drive temporal stability overall, and
together determine the sign of the stability-richness relationship.

@@

Ever-growing anthropogenic pressures have resulted in alteration of biodiversity
[@newbold_global_2015; @services_summary_2019], community changes
[@dornelas_assemblage_2014], and climate change will result in dramatic
environmental fluctuations [@lee_ipcc_2023]. How ecological communities will
respond to those threats is then a pressing, but challenging question
[@simmons_refocusing_2021; @yang_predictability_2019]. The temporal stability of
ecological communities is an emerging property that indeed results of the
complex interplay between anthropogenic pressures [@olivier_urbanization_2020;
@hautier_anthropogenic_2015; @yang_predictability_2019], community size
[@tilman_biodiversity_2006; @isbell_biodiversity_2009;
@olivier_urbanization_2020; @pennekamp_biodiversity_2018] and community
structure [@thebault_trophic_2005; @thebault_relationship_2006;
@thibaut_understanding_2013; @danet_species_2021;
@eschenbrenner_diversity_2023].

Understanding emerging temporal community stability (hereafter referred to as
community stability) requires scaling down the focus to population scale
[@doak_statistical_1998; @thibaut_understanding_2013; @zhao_biodiversity_2022].
Community stability can be partitioned into average population
stability and asynchrony [@loreau_species_2008; @thibaut_understanding_2013;
@zhao_biodiversity_2022], asynchrony measuring how much population
fluctuations’ are compensating at the community level. The question of how
global change will respectively affect population stability and asynchrony is
still open.

Asynchrony among species is the central pattern that drive the positive
community stability-diversity relation documented across many species assemblages
[@tilman_biodiversity_2006; @olivier_urbanization_2020;
@pennekamp_biodiversity_2018]. Asynchrony does not emerge by species richness
itself, but by its interplay with response diversity of species to perturbations
[@elmqvist_response_2003; @loreau_species_2008; @ives_stability_2000;
@ross_how_2023], and compensatory dynamics raising from species interactions.
Asynchrony can then stem from statistical averaging of independent species
fluctuations ["portfolio effects", @doak_statistical_1998;
@tilman_ecological_1999] or compensatory dynamics resulting from response
diversity and species interactions [@zhao_biodiversity_2022]. Among other
reasons, it might explain why reports of neutral or negative stability-richness
relationships are also common in empirical settings [@houlahan_negative_2018].
However, most of our knowledge comes from simplified community
structure, i.e. which does not account for the multiple trophic
levels constituting ecological communities [but see @danet_species_2021;
@thebault_trophic_2005; @ives_stability_2007; @eschenbrenner_diversity_2023],
thereby limiting our understanding of the drivers of both community stability
and stability-richness relationships in more complex communities.

The structure of community also shapes communities
[@macarthur_fluctuations_1955; @may_will_1972; @angelis_stability_1975;
@mccann_weak_1998; @vasseur_environmental_2007; @thebault_trophic_2005], such as
the strength [@may_will_1972; @mccann_weak_1998; @thebault_trophic_2005] and
distribution [@may_will_1972; @angelis_stability_1975; @mccann_weak_1998;
@mccann_reevaluating_1997] of species interactions determine population
stability and stability-richness relationship [two trophic level systems,
@thebault_trophic_2005; @ives_stability_2007], while the number of trophic
levels determine population stability [@shanafelt_stability_2018]. Community
structure also interact with response diversity to population stability in
diamond shaped food-webs [@ripa_food_2003; @vasseur_environmental_2007;
@ruokolainen_environmental_2013]. However, the work on community structure
lacked the complexity in terms of species richness or trophic levels, and
focused more on population stability than on asynchrony.

Then, the two body of work focusing on the effects of (1) species richness or
the effects of (2) community structure on community stability have been
developed in isolation. While (1) highlighted the importance of asynchrony
stemming response diversity ("insurance hypothesis"), (2) highlighted the
importance of population stability arising from species interactions (top-down
and bottom-up controls). Two keys advances allow to overcome that limitation:
the development of the Bioenergenetic model that enable to model arbitrary
complex communities [@yodzis_body_1992; @brose_allometric_2006] and the
development of statistical tools that enables to partition community stability
into population stability and asynchrony [@thibaut_understanding_2013;
@zhao_biodiversity_2022].

>>>>

- Putative titles:
  - Community stability is driven by biodiversity and response diversity in
    complex stochastic food-webs
  - Response diversity and food-web structure drive stability and
    stability-richness relationships in fluctuating environments

@@

The diversity of species responses to environmental stochasticity is key to
understand temporal community stability and stability-diversity relationships.
While we know quite well how response diversity and species richness drive
community stability in plant communities, whether these effects also operate
in complex food-webs remains an open question. Using a stochastic Bioenergenetic
model, we show that response diversity and species richness are also strong drivers
of overall community stability in complex food-webs. In contrast, we found that
interaction strength and connectance had little effects on temporal stability
overall. Negative stability-richness relationships always emerged in the absence
of response diversity, but we found that interaction strength, connectance,
average trophic level strongly interact with response diversity to determine the
sign of stability-diversity relationships. Our study disentangles the complex
effects of food-web structure on community stability and its pathway, showing
that response diversity is a mechanism that generate positive
stability-diversity relationships, beyond plant communities.

@@

Predicting how ecological communities respond ever-growing anthropogenic
pressures is a central focus in Ecology [@newbold_global_2015;
@services_summary_2019]. Most approaches so-far involve static environment and
simplified species assemblages, overlooking environmental stochasticity and the
complex structure of ecological communities. However, the effects of
stochasticity is much more than noise and lead to emerging properties at the
community level. The key mechanism is the way that species respond to
environmental stochasticity, the so called response diversity
[@elmqvist_response_2003; @loreau_species_2008; @ives_stability_2000;
@ross_how_2023]. Response diversity, even considered as pure random species
fluctuations, generate asynchrony in species fluctuations at the community
level. Coined as "statistical averaging", "portfolio effects" or "insurance
hypothesis", it is responsible for the positive relationships between temporal
stability of community and species richness in plant communities
[@doak_statistical_1998; @yachi_biodiversity_1999; @tilman_biodiversity_2006;
@zhao_biodiversity_2022]. But species responses to environmental fluctuations is
also constrained by species interactions, and so community structure
[@loreau_species_2008; @vasseur_environmental_2007; @thebault_trophic_2005;
@thebault_relationship_2006]. A pioneer example by @vasseur_environmental_2007,
in which they present a counter-intuitive example, where the lack of response
diversity can dampen species oscillations and then increase the temporal
stability in a four species diamond food-web [@mccann_weak_1998]. Despite its
overarching importance on community stability and stability-richness
relationships, the effects of stochasticity and response diversity in complex
communities, such as food-webs, remain mostly unknown.

Understanding emerging temporal community stability (hereafter referred to as
community stability) requires scaling down the focus to population scale
[@doak_statistical_1998; @thibaut_understanding_2013; @zhao_biodiversity_2022].
Community stability, measured as the inverse of the coefficient of variation,
can be partitioned into average population stability and asynchrony
[@thibaut_understanding_2013; @zhao_biodiversity_2022], which respectively
measures the amplitude of species fluctuations relative to their
biomass/abundance and how much species fluctuations’ are compensating each other
at the community level. Asynchrony can then stem from statistical averaging of
independent species fluctuations ["portfolio effects", @doak_statistical_1998;
@tilman_ecological_1999] or compensatory dynamics resulting from response
diversity to environmental fluctuations and species interactions
[@zhao_biodiversity_2022]. The body of theory of BEF, focusing mainly on plants,
showed that response diversity and species richness give rise to asynchrony and
thereby increase community stability, despite that species richness can decrease
population stability in the same time [@tilman_biodiversity_2006;
@houlahan_negative_2018]. On the other side, food-web theory focused more on
population stability. It shows that the strength [@may_will_1972;
@mccann_weak_1998; @thebault_trophic_2005] and distribution [@may_will_1972;
@angelis_stability_1975; @mccann_weak_1998; @mccann_reevaluating_1997] of
species interactions determine population stability ("complexity-stability"
theory) and stability-richness relationship [two trophic level systems,
@thebault_trophic_2005; @ives_stability_2007], while the number of trophic
levels determine population stability [@shanafelt_stability_2018]. Food-web
structure also interacts with response diversity to determine population
stability in diamond shaped food-webs [@ripa_food_2003;
@vasseur_environmental_2007; @ruokolainen_environmental_2013]. So far, food-web
theory overlooked the effects of food-web structure on response diversity and
asynchrony while BEF theory overlooked the complex effects of food-web structure
on population stability and asynchrony. Two keys advances allow to overcome that
limitation: the development of the Bioenergenetic model that enable to model
arbitrary complex communities [@yodzis_body_1992; @brose_allometric_2006] and
the development of statistical tools that enables to partition community
stability into population stability and asynchrony [@thibaut_understanding_2013;
@zhao_biodiversity_2022].
